Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD

### 2014-01-21 doc/papers/wpmvp14/paper.tex
Last update of the WPMVP paper
Real final version, as submitted to the ACM, version
Some code had to be reformatted to fit in the new page layout, some were
pruned due to an overgrowth of the paper.

% submit to https://sites.google.com/site/wpmvp2014/home
\documentclass[preprint]{sigplanconf}

@@

@@

%\doi{}

@@

%\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
%\copyrightyear{20yy} 
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

@@

\subsection{Interaction with Standard Algorithm}
Modern \cpp programming style based on Generic Programming usually leads to an
intensive use of various STL components like Iterators. \boostsimd{} provides
iterator adaptors that turn regular random access iterators into iterators
suitable for SIMD processing. These adaptors act as free functions taking
regular iterators as parameters and return iterators that output \texttt{pack}
whenever dereferenced. These iterators are then usable directly in usual STL
algorithms such as \texttt{transform} or \texttt{fold} (Listing \ref{stl_example}).

\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={SIMD Iterator with STL algorithm},label=stl_example]
std::vector<int,allocator<int> > v(N),r(N);

std::transform ( simd::begin(v.begin())
               , simd::end(v.end())
               , simd::begin(r.begin())
               , [](pack<int> const& p)
                 { 
                   return p*p; 
                 } 
              );
\end{lstlisting}

@@

\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={Polymorphic SIMD algorithm call},label=stl_poly]
std::vector<int,allocator<int> > v(128), r(128);

struct f
{
  template<class T>
  auto operator()(T const& p)
  -> decltype(p * p)
  {
    return p * p;
  }
};

simd::transform( v.begin(), v.end()
               , r.begin()
               , f() 
               );
\end{lstlisting}

@@

\begin{lstlisting}[language=c++, caption={Simplified Pythran Translation of a Polymorphic Function}, label={lst:simple-pythran-output}]
struct foo {
    template<class T>
    auto operator()(T&& x)
    -> decltype(3 * x)
    {
        return x * 3;
    }
};
\end{lstlisting}

@@

\begin{lstlisting}[language=python, label={lst:square-python}, caption={Python implementation of the square function.}]
def square(x):
  return x * x
\end{lstlisting}

\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={C++ meta-implementation of the square function.}]
struct square {
    template<class T>
    auto operator()(T const& x)
    -> decltype(x * x)
    {
      return x * x;
    }
};
\end{lstlisting}

@@

\begin{lstlisting}[language=python, label={lst:numpy-code}, caption={Sample \texttt{numpy} code.}]
import numpy as np
def rosen(x):
  return np.sum(100*(x[1:] - x[:-1] ** 2) ** 2
                + (1 - x[:-1]) ** 2)
\end{lstlisting}

@@

Experiments are run on a computer running a Linux kernel version 3.10. The
processors are Intel Core i7, so the AVX instruction set is available. Python
version is 2.7.5, Pythran version is the \texttt{HEAD} of the \texttt{wpmvp14}
branch of the official Pythran repository
\url{https://github.com/serge-sans-paille/pythran}. The back-end C++ compiler is
Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
performances.

@@

To evaluate the vectorization of \texttt{ndarray} point-to-point operations,
three expressions are used: \texttt{1. / a + a}, the example from
Listing~\ref{lst:numpy-code} and a computation of the point-to-point arc
distance  between random values. Again, all inputs are large randomized
\texttt{ndarray} of double precision floating point numbers.
Figure~\ref{fig:numpy-timings} summarizes the results. Because all computations
are done in native space, there is not much speedup when running them through
Pythran. The only benefits comes from the elimination of temporary arrays and
loop fusion. As a consequence, the more elements in the \texttt{numpy}
expression, the more Pythran accelerates the code. The vectorization speedups
follow the same process, as larger expression tend to increase the operation
over memory movement ratio.

@@

    \includegraphics[width=.5\textwidth]{numexpr}
    \includegraphics[width=.5\textwidth]{rosen}
    \includegraphics[width=.5\textwidth]{arc}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{fig:numpy-timings}

>>>>

% submit to https://sites.google.com/site/wpmvp2014/home
\documentclass[10pt]{sigplanconf}

@@

\conferenceinfo{WPMVP '14}{February 16, 2014, Orlando, FL, USA} 
\copyrightyear{2014} 
\copyrightdata{978-1-4503-2653-7/14/02}
\doi{2568058.2568060}

@@

@@

@@

\subsection{Interaction with Standard Algorithm}
Modern \cpp programming style based on Generic Programming usually leads to an
intensive use of various STL components like Iterators. \boostsimd{} provides
iterator adaptors that turn regular random access iterators into iterators
suitable for SIMD processing. These adaptors act as free functions taking
regular iterators as parameters and return iterators that output \texttt{pack}
whenever dereferenced. These iterators are then usable directly in usual STL
algorithms such as \texttt{transform} or \texttt{fold} (Listing \ref{lst:tl_example}).

\begin{lstlisting}[language=c++, label={lst:stl_example}, caption={SIMD Iterator with STL algorithm},label=stl_example]
std::vector<int,allocator<int>> v(N),r(N);
std::transform (
 simd::begin(v.begin()),simd::end(v.end()),
 simd::begin(r.begin()),
 [](pack<int> const& p) { return p*p; } 
);
\end{lstlisting}

@@

\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={Polymorphic SIMD algorithm call},label=stl_poly]
std::vector<int,allocator<int>> v(9), r(9);
struct f {
 template<class T>
 auto operator()(T const& p)
 -> decltype(p*p)
 { return p * p; }
};

simd::transform( v.begin(), v.end(),
                 r.begin(), f());
\end{lstlisting}

@@

\begin{lstlisting}[language=c++, caption={Simplified Pythran Translation of a Polymorphic Function}, label={lst:simple-pythran-output}]
struct foo {
 template<class T>
 auto operator()(T&& x) -> decltype(3*x)
 { return x * 3; }
};
\end{lstlisting}

@@

\begin{lstlisting}[language=python, label={lst:square-python}, caption={Python implementation of the square function.}]
square = lambda x: x * x
\end{lstlisting}

@@

\begin{lstlisting}[language=python, label={lst:numpy-code}, caption={Sample \texttt{numpy} code.}]
import numpy as np
def rosen(x):
  return np.sum(100*(x[1:]-x[:-1]**2)**2
                + (1 - x[:-1])**2)
\end{lstlisting}

@@

Experiments are run on a computer running a Linux kernel version 3.10. The
processors are Intel Core i7, so the AVX instruction set is available. Python
version is 2.7.5, Pythran version is the \texttt{HEAD} of the \texttt{wpmvp14}
branch of the official Pythran repository
\url{https://github.com/serge-sans-paille/pythran}. The back-end C++ compiler
is Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
performances. Using the compiler automatic vectorizer did not end up with any
speedup, most probably due to expression templates confusing the vectorizer.

@@

To evaluate the vectorization of \texttt{ndarray} point-to-point operations,
three expressions are used: \texttt{1. / a + a}, the example from
Listing~\ref{lst:numpy-code} and a computation of the point-to-point arc
distance  between random values. Again, all inputs are large randomized
\texttt{ndarray} of double precision floating point numbers.
Figures~\ref{fig:numpy-timings} and~\ref{fig:numpy-code-fs} summarize the
results. Because all computations are done in native space, there is not much
speedup when running them through Pythran. The only benefits comes from the
elimination of temporary arrays and loop fusion. As a consequence, the more
elements in the \texttt{numpy} expression, the more Pythran accelerates the
code. The vectorization speedups follow the same process, as larger expression
tend to increase the operation over memory movement ratio.

@@

    \includegraphics[width=.5\textwidth]{numexpr}
    %\includegraphics[width=.5\textwidth]{rosen}
    \includegraphics[width=.5\textwidth]{arc}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{fig:numpy-timings}
